---
title: "Data analysis report"
date: "`r Sys.Date()`"
author: "Alem Gusinac"

output:
  html_document:
    toc: true
    toc_float: true
---

<!--- Changes the report based on the browser changes-->
<!-- css formatting for text -->
<style type="text/css">
.main-container {
  width: 100%;
  max-width: 100%;
  min-width: 100%;
  margin-left: auto;
  margin-right: auto;
}
</style>
<!-- css formatting for figures -->
<style>
  img {
    max-width: 100%;
    height: auto;
  }
</style>
<!-- css formatting for tabulars -->
<style>
table {
  width: 100%;
  border-collapse: collapse;
}
</style>

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      echo = FALSE, 
                      results = "asis",
                      fig.width = 10,
                      fig.height = 7
                      ) 
```


```{r}
plot_tabs <- function(plots, method) {
  if (method == "matrix") {
    for(i in 1:nrow(plots)) {
      for(j in 1:ncol(plots)) {
        cat(paste0("\n\n### Plot ", i, ".", j, "\n"))
        print(plots[[i,j]])
        cat("\n")
      }
    }
  }
  if (method == "vector") {
    for (i in 1:length(plots)) {
      cat(paste0("\n\n### Plot ", i, "\n"))
      print(plots[[i]])
      cat("\n")
    }
  }
}

```

# Preprocessing step
Refer to published paper regarding pipeline\n

```{r}
preprocess_stats %>% 
  kableExtra::kbl(caption = "Table 1: Stats from cutadapt and DADA2 denoising") %>%
  kableExtra::kable_classic(full_width = F, html_font = "Calibri")

```

# Automated Analysis

## Rank stats
```{r}
rankstat_plot
```


## Alpha diversity {.tabset}
```{r}
plot_tabs(shannon_plots, method = "vector")
```

## Beta Diversity: Principal Coordinate Analysis (PCoA) of Weighted UniFrac distances {.tabset}
```{r}
plot_tabs(pcoa_plots, method = "matrix")
```

## Beta Diversity: NMDS {.tabset}
```{r}
plot_tabs(nmds_plots, method = "matrix")
```


## Composition {.tabset}
```{r}
plot_tabs(composition_plots, method = "matrix")
```

## Correlations {.tabset}
```{r}
plot_tabs(correlation_heatmap_plt, method = "vector")
```

## 2-fold heatmap {.tabset}
```{r}
# Unnaming the plot list and saving as a new list
final_plots <- list()
fold_headers <- c("boxplot", "barplot", "tile_plot", "rel_abun")
for (i in 2:length(heatmap_plots[[1]])) {
  final_plots[[i-1]] <- unname(heatmap_plots[[1]][fold_headers[i-1]])
}

# Separate function due to `[[1]]` being printed to stdout
for (i in 1:length(final_plots)) {
  cat(paste0("\n\n### Plot ", i, "\n"))
  print(final_plots[[i]][[1]])
  cat("\n")
}
```
<!-- ## Redundancy Analysis (RDA) {.tabset} -->
<!-- ```{r, fig.width=10, fig.height=10} -->
<!-- # par(mfrow=c(RANKSTAT_ncol, 2)) -->
<!-- # for (i in 1:RANKSTAT_ncol) { -->
<!-- #   RDA_plots[[i, 1]]() -->
<!-- #   RDA_plots[[i, 2]]() -->
<!-- # } -->

<!-- ``` -->
<!-- ## {-} -->