---
title: "main"
author: "A Gusinac"
date: "2024-03-18"
output: html_document
---

## Required Library's

```{r, knitr::opts_chunk$set, include = FALSE, warning=FALSE}
library("tidyverse")
library("biomformat")
library("phyloseq")
library("ape")
library("microViz")
library("decontam")
library("Biostrings")
library("patchwork")
library("vegan")
library("ComplexHeatmap")
```

## Load functions

```{r, include = FALSE, warning=FALSE}
source("../../../../general_scripts/R/99_project-functions.R")

# Utils
sourceDir(path = "../../../../general_scripts/R/utils")

```

## Specify data paths

```{r, eval=FALSE}
biom_filename <- "..//data//BiotaViz//relative-table-with-taxonomy.biom"
metadata_filename <- "..//data//metadata//metadata.tsv"
taxonomy_filename <- "..//data//classifier//vsearch//taxonomy//biotaviz_taxonomy.tsv"
tree_filename <- "..//data//phylogeny//phylogenetic_tree//tree.nwk"
refseq_filename <- "..//data//denoise//representative_sequences//unfiltered//sequences.fasta"
  
```

## Constructing physeq
```{r, eval=FALSE}
biom_data <- import_biom(BIOMfilename = biom_filename)
metadata <- import_qiime_sample_data(metadata_filename)
tree <- read_tree(tree_filename)
refseq <- readDNAStringSet(refseq_filename)

ps <- construct_ps(biom_data, metadata, tree, refseq)
#saveRDS(ps, "..//data//RDS//radboudumc_relative_phyloseq.rds")

```

# Systematic comparison
## Loading phyloseq objects

```{r, include = FALSE, warning=FALSE}
radboud_ps_abs <- readRDS("..//data//RDS//radboudumc_absolute_phyloseq.rds")

```

## Standard Data wrangling
```{r}
# data wrangling
#------------------------------------------------------------------------------#
radboud_ps_abs_bac <- radboud_ps_abs %>% 
  subset_taxa(Domain == "Bacteria") 

# normalize by bacterial domain
radboud_ps_rel_bac_norm <- radboud_ps_abs_bac %>% 
  transform_sample_counts(function(x) x / sum(x))

```

## Rankstat
### Classified ASVs per taxonomic rank

```{r, warning=FALSE}
# subsetting only for positive samples
radboud_ps_positive <- radboud_ps_abs %>% 
  subset_samples(RANKSTAT_treatment != "control") %>% 
  subset_taxa(Domain == "Bacteria")

maastricht_proportion_classified <- plot_proportion_classified(maastricht_ps)
radboud_proportion_classified <- plot_proportion_classified(radboud_ps_positive)

(maastricht_proportion_classified + radboud_proportion_classified) + 
  plot_annotation(title = "Maastricht versus Radboud FFPE samples",
                  subtitle = "Percentage of the number of ASVs classified at each rank") +
  plot_layout(guides = "collect",
              axis_titles = "collect")

```

## Proportion of classified ASVs per taxonomic rank for each sample

```{r, warning=FALSE, fig.width=10, fig.height=5}
maastricht_norm_comp <- proportion_samples(maastricht_ps, sample_format=FALSE)
radboud_norm_comp <- proportion_samples(radboud_ps_positive, sample_format=TRUE) +
  theme(axis.text.y = element_blank())

comb <- (maastricht_norm_comp + radboud_norm_comp) + 
  plot_annotation(title = "Maastricht versus Radboud FFPE samples",
       subtitle = "ASVs normalised per sample showing top 15 relative abundant taxa",
       caption = paste("Spearman correlation between two phyloseqs is ")) +
  theme(
    text=element_text(size=20),
    plot.title = element_text(size=25)
  )

ggsave(filename = "..//results//Proportion_classified_combination.png",
       plot = comb,
       width = 30,
       height = 18,
       limitsize = FALSE)

```

### Spearman correlation (non-paramatic test)

```{r, warning=FALSE}
# Spearman correlation
sample_names(radboud_ps_positive) <- sample_names(radboud_ps_positive) %>%
  str_extract("(^[^_]*_[^_]*_?\\d)|(C\\d?)")

norm_maastricht <- transform_sample_counts(maastricht_ps, function(ASV) ASV / sum(ASV))
norm_radboud <- transform_sample_counts(radboud_ps_positive, function(ASV) ASV / sum(ASV))


top15_taxa_maastricht <- norm_maastricht %>% 
  tax_fix(anon_unique = FALSE,
          verbose = FALSE,
          unknowns = c("uncultured")) %>% 
  tax_glom(taxrank = "Genus", NArm = FALSE) %>% 
  tax_top(n = 15)

top15_taxa_radboud <- norm_radboud %>% 
  tax_fix(anon_unique = FALSE,
          verbose = FALSE,
          unknowns = c("uncultured")) %>% 
  tax_glom(taxrank = "Genus", NArm = FALSE) %>% 
  tax_top(n = 15)

maastricht_abun <- t(otu_table(norm_maastricht))[top15_taxa_maastricht, ] %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything())

radboud_abun <- otu_table(norm_radboud)[top15_taxa_radboud, ] %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything())

spearman_cor <- cor.test(maastricht_abun$value, radboud_abun$value, method = "spearman")
kruskal_test <- kruskal.test(maastricht_abun$value ~ radboud_abun$value)

```

### Distribution of ASVs, relative and absolute in treatment groups

```{r, warning=FALSE}
# Subset by taxa: Change Subset_taxa by desired filter rank
#------------------------------------------------------------------------------#
ps <- radboud_ps_abs_bac %>% 
  subset_taxa(Genus == "Pseudomonas")

# Computes sum of ASVs per sample
target_ps <- ps %>% 
  ps_mutate(nASVs = sample_sums(ps))

# Defines unique groups
unique_groups <- unique(sample_data(target_ps)$RANKSTAT_treatment) 

# Subsets by RANKSTAT_treatment into ASVs and Phyloseq objects
ASVs_list <- list()
ps_list <- list()

for (variable in unique_groups) {
  ps_subset <- subset_samples(target_ps, RANKSTAT_treatment == {{variable}}) %>% removeZeros()
  ASV_subset <- ps_subset %>% removeZeros() %>% taxa_names()
  
  # store in list
  ps_list[[variable]] <- ps_subset
  ASVs_list[[variable]] <- ASV_subset
}

# Finds unique ASVs pairwise between treatment groups
group_combi <- combn(x = unique_groups, m = 2)
ASVs_diff_list <- list()
for ( i in seq(ncol(group_combi))) {
  pair <- group_combi[, i]
  ASVs_diff <- setdiff(ASVs_list[[pair[2]]], ASVs_list[[pair[1]]])
  ASVs_diff_list[[paste0(pair[2],"_vs_",pair[1])]] <- ASVs_diff
}

# Plots original relative abundance and absolute abundance side by side
for (var in unique_groups) {
  plt <- relative_and_absolute_plot(ps_list[[var]], tax_target = "Species", sample_format = TRUE)
  
  ggsave(filename = paste0("../results/Genus_", var, ".png"),
       plot = plt,
       width = 20,
       height = 15,
       limitsize = FALSE)
}

```
### Correcting treatment groups by different ASVs
```{r}
ps_healthy_filt <- ASVs_filter(ps_healthy, ASVs_ht_vs_ct)
ps_tumor_filt <- ASVs_filter(ps_tumor, ASVs_tu_vs_ct)

ps_healthy_filt_stat <- ps_healthy_filt %>% 
  ps_mutate(nASVs = sample_sums(ps_healthy_filt))

ps_tumor_filt_stat <- ps_tumor_filt %>% 
  ps_mutate(nASVs = sample_sums(ps_tumor_filt))

relative_and_absolute_plot(ps_healthy_filt_stat, sample_format = TRUE)
relative_and_absolute_plot(ps_tumor_filt_stat, sample_format = TRUE)


```


### Distribution of treatment groups in Phylum (neat)
```{r}
radboud_ps_abs %>% 
  subset_taxa(Domain == "Bacteria") %>% 
  tax_fix() %>% 
  merge_samples(group = "RANKSTAT_treatment") %>% 
  comp_barplot(
    tax_level = "Phylum", n_taxa = 15,
    sample_order = unique_groups,
    bar_width = 0.8
  ) +
  coord_flip() + 
  labs(x = NULL, y = NULL)
```

### Heatmap correlation plot

```{r}
cor_heatmap_plot(radboud_ps_abs, tax_target = "Genus")
```


## Alpha diversity 
### Shannon index graph: data from nextflow run
```{r}
df_shannon <- read_csv(file = "../data/diversity_metrics/alpha_rarefaction/shannon.csv")

df_shannon %>% 
  pivot_longer(cols = starts_with("depth-"),
               names_to = "iters",
               values_to = "alpha_div") %>% 
  ggplot(mapping = aes(x = RANKSTAT_treatment,
                       y = alpha_div,
                       fill = RANKSTAT_treatment)) +
  geom_violin(width = 1.4) +
  geom_boxplot(width = 0.1, 
               color = "black",
               alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Alpha diversity",
       x = "sample groups",
       y = "Shannon Index")

```

### Working from observed counts: data from nextflow run
```{r}
df_obs_features <- read_csv(file = "../data/diversity_metrics/alpha_rarefaction/observed_features.csv")

# replace NAs by zero
df_obs_clean <- df_obs_features %>% 
  select(starts_with("depth")) %>% 
  list() %>% 
  replace_na(0) %>% 
  as.data.frame() %>% 
  as.matrix()


df_total_clean <- df_obs_clean %>% 
  rowwise(samples) %>% 
  select(-samples) %>% 
  mutate(total = sum(c_across(where(is.numeric)))) 

shannon_vec <- vegan::diversity(x = df_obs_clean, 
                 index = "shannon")

df_samples <- data.frame(
  samples = df_obs_features$`sample-id`,
  groups = df_obs_features$RANKSTAT_treatment,
  shannon = shannon_vec
)

df_samples %>% 
  ggplot(mapping = aes(x = groups,
                       y = shannon)) +
  geom_violin(width = 1.4) +
  geom_boxplot(width = 0.1, 
               color = "black",
               alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Alpha diversity",
       x = "sample groups",
       y = "Shannon-Wiener Index (H')")

```

## Species, Richness, Eveness statistics from biom format
```{r}
otu_tab <- ps2otu(ps = radboud_ps_abs_bac %>% 
                    subset_samples(RANKSTAT_treatment != "control") %>% 
                    removeZeros(), 
                  tax_target = "Genus")

# replace positives by 1
otu_tab[otu_tab > 1] <- 1

otu_tab <- otu_tab %>% t()
group <- sample_data(radboud_ps_abs_bac)$RANKSTAT_treatment


df_alpha <- data.frame(
  samples = colnames(otu_tab),
  groups = group[group != "control"],
  species_n = vegan::specnumber(otu_tab, MARGIN = 2),
  shannon = vegan::diversity(otu_tab, index = "shannon", MARGIN = 2),
  simpson = vegan::diversity(otu_tab, index = "simpson", MARGIN = 2)
)

df_alpha %>% 
  ggplot(mapping = aes(x = species_n)) +
  geom_line(aes(y = shannon, color = "black")) +
  geom_line(aes(y = simpson, color = "red")) +
  scale_y_continuous(
    sec.axis = sec_axis(~./2, name="Second Axis")
  )
```


## Principal Coordinate Analysis with Unifrac
```{r}
# normalize by bacterial domain
#------------------------------------------------------------------------------#
radboud_ps_rel_bac_norm <- radboud_ps_abs_bac %>% 
  transform_sample_counts(function(x) x / sum(x))

comb_plot <- wunifrac_pcoa(radboud_ps_rel_bac_norm)
ggsave("../results/test.png",
       plot = comb_plot,
       width = 10,
       height = 10)

```

## Heatmap log transformed without Pseudomonas
```{r}
## TO DO: Rewrite function

ps_fold_heatmap <- function(ps, threshold = 0.2, taxa_rank = "Genus", negative_ct="healthy", positive_ct="tumor") {
  # Fetch otus and filter out less representative Genera
  otu_tab <- ps2otu(ps, tax_target = taxa_rank) %>% t()
  otu_tab.filt <- otu_tab[rowSums(otu_tab) > threshold, ]
  
  # log2 scale
  otu_tab.trans <- otu_tab.filt %>%
    as.matrix() %>% 
    as("sparseMatrix") %>% 
    log2() %>% 
    as.matrix()
  
  # replace infinite by zero
  otu_tab.trans[!is.finite(otu_tab.trans)] <- 0
  
  # Creates dataframe & orders taxa abundances
  df <- as.data.frame(otu_tab.trans)
  df$Taxa <- rownames(df) 
  df$Total <- rowSums(df[, -ncol(df)])
  df <- df[order(-df$Total, decreasing=TRUE), ]
  
  # Pivot longer table by specified id.vars
  df_melt <- reshape2::melt(df, id.vars = c("Taxa", "Total"))
  
  # Removes uncultured genera, cleans sample names
  df_final <- df_melt %>% 
    filter(!grepl("uncultured", Genera)) %>% 
    mutate(
      # Specific for data, atm hardcoded but should be automated!
      group = ifelse(base::grepl("T_", variable), "tumor",
                     ifelse(base::grepl("H_", variable), "healthy", NA)),
      sample.id = sub("(-?\\w{1}_\\d+)$", "" ,variable)
    ) %>% 
    group_by(sample.id) %>% 
    filter(any(group == negative_ct) & any(group == positive_ct))
  
  # Computes log2(tumor) - log2(healthy)
  df_diff <- df_final %>% 
    group_by(sample.id, Genera) %>% 
    summarise(diff = value[group == positive_ct] - value[group == negative_ct])
  
  # Generate heatmap plot with df_diff data
  heatmap_plot <- df_final %>% 
    left_join(df_diff, by = c("Taxa", "sample.id")) %>% 
    ggplot(mapping = aes(x = as.factor(sample.id),
                         y = Genera,
                         fill = diff)) +
    geom_tile() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.title.y = element_blank(),
          strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF")) +
    scale_fill_gradient2(name = "log2(tumor / healthy) fold",
                        low = "blue",
                        mid = "white",
                        high = "red") +
    scale_y_discrete(limits = rev(levels(as.factor(df_final$Genera)))) +
    labs(x = "Samples", 
         y = "Taxa")
  
  # Fetch otu table for boxplot and reshapes into long table
  otu_tab.tumor <- as.data.frame(otu_tab.filt)
  otu_tab.tumor$Genera <- rownames(otu_tab.filt)
  otu_tab.tumor_ <- reshape2::melt(otu_tab.tumor, id.vars = c("Taxa"))
  
  rel_abun_plot <- otu_tab.tumor_ %>% 
    ggplot(mapping = aes(x = value,
                         y = Genera)) +
    geom_boxplot() +
    theme_bw() +
    theme(axis.title.y = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank()) +
    scale_x_continuous(breaks=seq(0, 1, 0.5)) +
    scale_y_discrete(limits = rev(levels(as.factor(df_final$Genera)))) +
    labs(x = "Relative Abundance")
  
  comb_plot <- (heatmap_plot + rel_abun_plot) +
    plot_layout(guides = "collect",
                axes = "collect",
                widths = c(5,2)) +
    plot_annotation(title = "Microbiome of Breast Carcinoma patients",
                    subtitle = "")
  
  return(comb_plot)
} 

ps <- radboud_ps_abs_bac %>% 
  subset_samples(RANKSTAT_treatment != "control") %>% 
  subset_taxa(Genus != "Pseudomonas") %>% 
  transform_sample_counts(function(x) x / sum(x)) %>% 
  removeZeros()

combined_plot <- ps_fold_heatmap(ps)

# ggsave(filename = "../results/heatmap_2fold_change.png",
#        plot = comb_plot,
#        width = 10,
#        height = 10,
#        limitsize = FALSE)

```