---
title: "Data Analysis"
author: "RTC Bioinformatics"
date: "2024-03-18"
output: pdf_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE) 
```

## Required Library's

```{r, warning=FALSE}
library("foreach")
library("tidyr")
library("ggplot2")
library("magrittr")
library("Matrix")
```

## Load functions

```{r, warning = FALSE}
source("C:/Users/Z289224/Documents/general_scripts/R/OmicFlow/R/99_project-functions.R")

# Utils
sourceDir(path = "C:/Users/Z289224/Documents/general_scripts/R/OmicFlow/R/utils_v2")

```
## Specify data paths

## Constructing metataxonomics class
```{r, eval=FALSE}
current_path <- getwd()
taxa_class <- metataxonomics$new(metaData = paste0(current_path, "/data/metadata.tsv"),
                           biomData = paste0(current_path, "/data/biom_with_taxonomy.biom"),
                           treeData = paste0(current_path, "/data/rooted_tree.newick"))

```

# General analysis flow
## Rankstat
### Classified ASVs per taxonomic rank

```{r}
rankstat_plot <- taxa_class$rankstat()

```

## Proportion of classified ASVs per taxonomic rank for each sample

```{r}
taxa_class$feature_subset(Domain == "Bacteria")
result <- taxa_class$composition(feature_rank = "Genus",
                                 feature_filter = c("uncultured"),
                                 feature_top = 10)
taxa_class$reset()

composition_plot(data = result$data,
                 palette = result$palette,
                 feature_rank = "Genus")

```

## Alpha diversity 
### Shannon index graph: data from nextflow run
```{r}
df_shannon <- data.table::fread(paste0(current_path, "/data/shannon.csv"))

taxa_class$feature_subset(Domain == "Bacteria")
shannon_plot <- taxa_class$shannon(df_shannon = df_shannon,
                                   col_name = "RANKSTAT_treatment")
taxa_class$reset()

```


## Principal Coordinate Analysis with Weighted Unifrac distances
```{r}
# normalized by bacterial domain
taxa_class$feature_subset(Domain == "Bacteria")
pcoa_plots <- taxa_class$ordination(metric = "unifrac",
                                    method = "pcoa",
                                    group_by = "RANKSTAT_treatment", 
                                    weighted = TRUE,
                                    parallel = TRUE,
                                    normalize = TRUE)
taxa_class$reset()

ggsave("results/pcoa_test.png",
       plot = comb_plot,
       width = 15,
       height = 5)

```

```{r}
# Create Linear regressin (incl. RDA, CCA etc.)
# Nested function #1: Find dimensions for 80% variance explained

  


library("ggplot2")
library("magrittr")
library("vegan")

data("dune")
data("dune.env")


triplot(counts = dune,
        metadata = dune.env,
        metadata.col = "Management")


```

## Heatmap log2 fold change [IMPROVING VISUALIZATION]
```{r}
  # Required parameters
  feature_rank <- "Genus"
  sample.id <- "SAMPLE-ID"
  paired <- FALSE
  paired.id <- "patient"
  feature_filter <- c("uncultured")
  feature_top <- NA
  condition.group <- "response"
  condition_A <- c("response")
  condition_B <- c("non-responders")
  normalize <- TRUE
  pvalue.threshold <- 0.05
  foldchange.threshold <- 0.6
  
  taxa$reset()
  taxa$feature_subset(Domain == "Bacteria")
  
  # Final output
  plot_list <- list(
        data = NULL,
        box_plot = NULL,
        volcano_plot = NULL,
        tile_plot = NULL,
        rel_abun = NULL,
        pvalues = NULL,
        volcano = NULL
      )
      
      # Copies object to prevent modification of tools class components
      counts <- data.table::copy(self$countData)
      metadata <- data.table::copy(self$metaData)
      features <- data.table::copy(self$featureData)
      
      #------#
      # Main #
      #------#
      
      # Subset samples by conditions
      # self$sample_subset(self$metaData[[condition.group]] %in% c(condition_A, condition_B))
      
      if (normalize) {
        taxa$transform(function(x) x / sum(x))
      }
      
      # Agglomerate taxa by feature rank and filter unwanted taxa
      taxa$feature_glom(feature_rank = feature_rank, feature_filter = feature_filter)
      
      # Check how many features to select (depended if volcano is desired)
      if (!is.na(feature_top)) {
        feature_top <- feature_top
      } else {
        feature_top <- nrow(taxa$featureData)
      }
      
      # Extract relative abundance
      rel_abun <- rowMeans(taxa$countData[1:feature_top, .SD, .SDcols = colnames(taxa$countData)])
      
      # Creates long table of relative abundance
      dt <- taxa$countData[, (feature_rank) := taxa$featureData[[feature_rank]]]
      stats_dt <- base::merge(data.table::melt(dt,
                                               measure.vars = colnames(dt)[!grepl(feature_rank, colnames(dt))],
                                               variable.name = sample.id, 
                                               value.name = "values"),
                              taxa$metaData[, .SD, .SDcols = c(sample.id, condition.group)],
                              by = sample.id)
      
      # Create row_sums
      dt[, row_sum := rowSums(.SD), .SDcols = !c(feature_rank)]
      
      # Orders by row_sum in descending order
      taxa$countData <- data.table::setorder(dt, -row_sum)[1:feature_top, .SD, .SDcols = !c("row_sum")]
      features <- taxa$countData[[ feature_rank ]]
      taxa$countData <- taxa$countData[, .SD, .SDcols = !c(feature_rank)]
      
      # Log2 transform taxa
      taxa$transform(log2)
      
      # Subset by top features
      stats_dt <- stats_dt[stats_dt[[feature_rank]] %in% features]
      dt <- taxa$countData[, (feature_rank) := features]
      
      # Compute 2-fold expression based on (un)paired samples
      # Computes on equation oflog2(A) - log2(B)
      # Supports multiple inputs for A and B.
      # For example A = T1, T2 and B = H1, H2
      if (paired == TRUE) {
        # sorting of metadata
        condition.labels <- data.table::setorderv(taxa$metaData, 
                                                  cols = c(sample.id, paired.id, condition.group))[[ condition.group ]]
        # paired samples
        DFE <- paired_fold(dt = dt,
                           sample.id = sample.id,
                           paired.id = paired.id,
                           condition_A = condition_A,
                           condition_B = condition_B,
                           unique.id = unique(taxa$metaData[[ paired.id ]]),
                           condition_labels = condition.labels,
                           feature_rank = feature_rank,
                           cpus = cpus)
        # Save data
        plot_list$data <- DFE
        
      } else if (paired == FALSE) {
        # sorting of metadata
        condition.labels <- data.table::setorderv(taxa$metaData,
                                                  cols = c(sample.id, condition.group))[[ condition.group ]]
        # unpaired samples
        DFE <- unpaired_fold(dt = dt,
                             sample.id = sample.id,
                             condition_A = condition_A,
                             condition_B = condition_B,
                             condition_labels = condition.labels,
                             feature_rank = feature_rank,
                             cpus = cpus)
        # Save data
        plot_list$data <- DFE$data
        
      } else {
        stop("paired can only be TRUE or FALSE, check your input.")
      }
      
      # Generate heatmap plot with df_diff data
      if (paired == TRUE) {
        # Adds size to paired heatmap 
        add_columns <- unique(taxa$metaData[, .SD, .SDcols = c(sample.id, paired.id)])
        
        merged_data <- base::merge(
          stats_dt,
          add_columns,
          by = sample.id,
          all.x = TRUE
        )
        
        # Subset merged data
        subset_merged <- merged_data[, .SD, .SDcols = c(paired.id, feature_rank, "values")]
        colnames(subset_merged) <- c("SAMPLE-ID", feature_rank, "values")
        
        # Second merge
        final_merge <- base::merge(
          x = DFE,
          y = subset_merged,
          by = c("SAMPLE-ID", "Genus"),
          all.x = TRUE
        )
      
        # Check if multiple diff_ are present 
        grouped_dt <- final_merge %>% 
          dplyr::group_by(`SAMPLE-ID`, Genus) %>% 
          dplyr::summarise(mean_values = mean(values, na.rum = TRUE),
                           diff_1 = mean(diff_1, na.rm = TRUE)) %>% 
          dplyr::ungroup()
        
        
        # Generating heatmap plot based on paired boolean
        n_diff_columns <-  sum(grepl("^diff_", colnames(DFE)))
    
        # Generate heatmap plot with df_diff data
        heatmap_plot <- grouped_dt %>% 
          ggplot(mapping = aes(x = base::get(sample.id, DFE),
                               y = base::get(feature_rank, DFE)))
        
        # If there is only one column uses default settings
        if (n_diff_columns == 1) {
          heatmap_plot <- heatmap_plot +
            geom_point(aes(size = mean_values, fill = diff_1), shape = 21)
        } else {
          # Adds geom_tile for number of diff_columns
          for (i in 1:n_diff_columns) {
            if (i == 1) {
              heatmap_plot <- heatmap_plot +
                geom_point(aes(size = mean_values, fill = !!sym(paste0("diff_", i))), shape = 21)
            } else {
              heatmap_plot <- heatmap_plot +
                geom_point(aes(size = mean_values, fill = !!sym(paste0("diff_", i))), 
                           shape = 21,
                           position = position_nudge(x = 0.5))
            }
            
          }
        }
        # Finishes heatmap plot
        plot_list$tile_plot <- heatmap_plot +
          theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
              axis.text.y = element_text(size=12),
              axis.text = element_text(size=12),
              text = element_text(size=12),
              legend.text = element_text(size=12),
              legend.title = element_text(size=14),
              strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF")) +
        scale_y_discrete(limits = rev(levels(as.factor(grouped_dt[["Genus"]])))) +
        scale_fill_gradient2(name = paste0("log2( A / B )"),
                             low = "blue",
                             mid = "white",
                             high = "red",
                             na.value = "grey80") +
        scale_size_continuous(name = "Mean Rel. Abun. (%)", labels = scales::label_number(accuracy = 0.01)) +
        labs(x = NULL, 
             y = NULL)  
        
        
        # Creates boxplot from relative abundances
        plot_list$rel_abun <- stats_dt %>% 
          ggplot(mapping = aes(x = values,
                               y = .data[[ feature_rank ]])) +
          facet_wrap(~.data[[condition.group]], ncol = length(condition_A) + length(condition_B)) +
          geom_boxplot() +
          theme_bw() +
          theme(text=element_text(size=12),
                axis.text.x = element_text(angle = 45, hjust = 1),
                axis.title.y = element_blank(),
                axis.text.y = element_blank(),
                axis.ticks.y = element_blank(),
                panel.spacing.x = unit(1, "lines")) +
          scale_x_continuous(trans = scales::log_trans()) +
          labs(x = "Log10( Rel. Abun. )")
      } else {
        # Store pvalues and volcano dt
        plot_list$volcano <- DFE$volcano
        plot_list$pvalues <- DFE$pvalues
        
        #----------------------#
        # Volcano plot         #
        #----------------------#
        # Create merged dt for volcano with mean foldchange and rel. abundance
        volcano_merged <- DFE$volcano[, "rel_abun" := rel_abun]
        
        # Create & save volcano plot
        plot_list$volcano_plot <- volcano_plot(dt = volcano_merged, 
                                               feature_rank = feature_rank,
                                               pvalue.threshold = pvalue.threshold,
                                               logfold.threshold = foldchange.threshold)
        
        # Subset significant groups
        significant_subset <- volcano_merged[volcano_merged[, pvalue < pvalue.threshold & (foldchange > foldchange.threshold | foldchange < -foldchange.threshold)], ]
        significant_final <- DFE$data[DFE$data[[feature_rank]] %in% significant_subset[[feature_rank]], ]
        
        # Use significant values to create boxplot
        plt <- significant_final %>% 
          ggplot(aes(y = .data[[ "diff_1" ]],
                     x = .data[[feature_rank]],
                     color = .data[["diff_1"]])) +
          gghalves::geom_half_boxplot() +
          gghalves::geom_half_point() +
          coord_flip() +
          theme_bw() +
          theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
                axis.text.y = element_text(size=12),
                axis.text = element_text(size=12),
                text = element_text(size=12),
                legend.text = element_text(size=12),
                legend.title = element_text(size=14),
                legend.position = "none",
                axis.title.y = element_blank(),
                strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF")) +
          scale_color_gradient2(name = paste0("log2( A / B )"),
                               low = "blue",
                               mid = "white",
                               high = "red",
                               na.value = "grey80") +
          scale_x_discrete(limits = rev(levels(as.factor(significant_final[[ feature_rank ]])))) +
          labs(x = NULL,
               y = NULL)
          
          
ggsave("test_gghalves.png", plot = plt) 
        
        # Use significant values to create rel abun plot
        
                # Generating heatmap plot based on paired boolean
        n_diff_columns <-  sum(grepl("^diff_", colnames(DFE$data)))
        # Creates boxplot and barplot for unpaired samples
        for (k in c("boxplot")) {
          plot_list[[k]] <- patchwork::wrap_plots(
            lapply(1:n_diff_columns,
                   function(i) fold_plot(dt = DFE$data, 
                                         X = paste0("diff_", i), 
                                         Y = feature_rank,
                                         title = paste0("Log2 ( ", condition_A[i], " / ", condition_B[i], " )"), 
                                         method = k, 
                                         taxa_labels = i == 1, 
                                         pvalues = DFE$pvalues)),
            ncol = n_diff_columns,
            nrow = 1)
        }
      }
      
      # Restores tools class components
      self$countData <- counts
      self$featureData <- features
      self$metaData <- metadata

```
