---
title: "Data Analysis"
author: "RTC Bioinformatics"
date: "2024-03-18"
output: pdf_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE) 
```

## Required Library's

```{r, warning=FALSE}
library("OmicFlow")
```

## Load functions

```{r, warning = FALSE}
source("C:/Users/Z289224/Documents/general_scripts/R/OmicFlow/R/99_project-functions.R")

# Utils
sourceDir(path = "C:/Users/Z289224/Documents/general_scripts/R/OmicFlow/R/utils_v2")

```
## Specify data paths

## Constructing metataxonomics class
```{r, eval=FALSE}
current_path <- getwd()
taxa <- metataxonomics$new(metaData = paste0(current_path, "/data/metadata.tsv"),
                           biomData = paste0(current_path, "/data/biom_with_taxonomy.biom"),
                           treeData = paste0(current_path, "/data/rooted_tree.newick"))

taxa_class$feature_subset(Domain == "Bacteria")
result <- taxa_class$composition(feature_rank = "Genus",
                                 feature_filter = c("uncultured"),
                                 feature_top = 10)

composition_plot(data = result$data,
                 palette = result$palette,
                 feature_rank = "Genus")

```

# General analysis flow
## Rankstat
### Classified ASVs per taxonomic rank

```{r}
rankstat_plot <- taxa_class$rankstat()

```

## Proportion of classified ASVs per taxonomic rank for each sample

```{r}
taxa_class$feature_subset(Domain == "Bacteria")
result <- taxa_class$composition(feature_rank = "Genus",
                                 feature_filter = c("uncultured"),
                                 feature_top = 10)
taxa_class$reset()

composition_plot(data = result$data,
                 palette = result$palette,
                 feature_rank = "Genus")

```

## Alpha diversity 
### Shannon index graph: data from nextflow run
```{r}
df_shannon <- data.table::fread(paste0(current_path, "/data/shannon.csv"))

# Pivot into long table
shannon_long <- data.table::melt(data = df_shannon,
                                 measure.vars = colnames(df_shannon)[grepl("depth-", colnames(df_shannon))], 
                                 variable.name = "iters",
                                 variable.factor = FALSE,
                                 value.name = "alpha_div")
# Corrects colnames
colnames(shannon_long) <- c("SAMPLE-ID", "iters", "alpha_div")
# Adds new column
shannon_final <- base::merge(shannon_long, 
                             taxa$metaData[, .SD, .SDcols = c("SAMPLE-ID", col_name)], 
                             by = "SAMPLE-ID", 
                             all.x = TRUE)

taxa$shannon(custom_div = shannon_final)

taxa_class$feature_subset(Domain == "Bacteria")
shannon_plot <- taxa_class$shannon(df_shannon = df_shannon,
                                   col_name = "RANKSTAT_treatment")
taxa_class$reset()

```


## Principal Coordinate Analysis with Weighted Unifrac distances
```{r}
# normalized by bacterial domain
taxa_class$feature_subset(Domain == "Bacteria")
pcoa_plots <- taxa_class$ordination(metric = "unifrac",
                                    method = "pcoa",
                                    group_by = "RANKSTAT_treatment", 
                                    weighted = TRUE,
                                    parallel = TRUE,
                                    normalize = TRUE)
taxa_class$reset()

ggsave("results/pcoa_test.png",
       plot = comb_plot,
       width = 15,
       height = 5)

```

## TREE with extra layers [IMPROVING VISUALIZATION]
```{r}
library(ggplot2)
library(ggtree)

x <- rtree(20)
df <- data.frame(row.names = x$tip.label,
                 cor = rnorm(20, mean = 0, sd = 0.5),
                 rel = rnorm(20, mean = 0.5, sd = 0.5),
                 fold = rnorm(20, mean = 0, sd = 2))
tree <- dplyr::full_join(x, df, by = "label")

p <- x %>% 
  ggtree(layout = "circular", branch.length = "none") +
  geom_tiplab(size = 2.5)

gheatmap(p, df[, "cor", drop=FALSE], 
         offset = 0.1, 
         width = 0.2, 
         colnames_position = "top",
         colnames_angle = 90,
         hjust =0,
         font.size = 2.5,
         low = "blue", high = "red", color="black")
  new_scale_fill()
  gheatmap(p, df[, "rel", drop=FALSE], 
         offset = 2, 
         width = 0.2, 
         colnames_position = "top",
         colnames_angle = 90,
         hjust =0,
         font.size = 2.5,
         color = "gray30") +
  new_scale_fill()
  


ggsave("test.png", p2)
p2

```

# Clustering methods
```{r}



```

