---
title: "Data Analysis"
author: "RTC Bioinformatics"
date: "2024-03-18"
output: pdf_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = TRUE, message = TRUE, echo = FALSE) 
```

## Required Library's

```{r, warning=FALSE}
library("Matrix")
library("ggplot2")
library("patchwork")
devtools::load_all()

taxa <- metagenomics$new(metaData = "../../../Pathology/projects/PRELEDE/data/metagenomics/metadata.tsv",
                         biomData = "../../../Pathology/projects/PRELEDE/data/metagenomics/contig_annotation/CAT_with_taxonomy.biom")

dt <- sparse_to_dtable(taxa$countData)
OTU_ID <- taxa$featureData$V1

dt1 <- cbind(OTU_ID, dt)
colnames(dt1) <- c("#OTU ID", colnames(dt1)[-1])

data.table::fwrite(x = dt1,
                   file = "abundance.tsv", sep = "\t")
```

```{r}
bench::mark(
taxa <- metagenomics$new(biomData = "../../../RTC/projects/lowlands_16s/results/biom_with_taxonomy.biom",
                         metaData = "../../../RTC/projects/lowlands_16s/results/metadata.tsv")
)
# taxa <- metagenomics$new(biomData = "../../../Pathology/projects/FFPE_breast_microbiome/data/biom_with_taxonomy.biom",
#                          metaData = "../../../Pathology/projects/FFPE_breast_microbiome/data/metadata.tsv")

taxa$feature_subset(Kingdom == "Bacteria")

res <- taxa$composition(
  feature_rank = "Genus",
  feature_filter = c("uncultured"),
  col_name = "attraction_score",
  feature_top = 10
)

comp_plt <- composition_plot(
  data = res$data,
  palette = res$palette,
  feature_rank = "Genus"
  )

ggsave(
  filename = "../../../RTC/projects/lowlands_16s/results/lowlands_composition_attractionscore.png",
  plot = comp_plt,
  scaling = 1,
  width = 7,
  height = 6,
  dpi = 400
  )

data.table::fwrite(
  x = res$data,
  file = "../../../RTC/projects/lowlands_16s/results/lowlands_composition.tsv",
  sep = "\t"
)

dfe <- taxa$DFE(
  feature_rank = "Genus",
  feature_filter = c("uncultured"),
  paired = FALSE,
  condition.group = "response",
  condition_A = c("high-responders"),
  condition_B = c("adjuvant")
)

dfe$volcano_plot[[1]]
View(dfe$data)

plt <- volcano_plot(
  data = dfe$data,
  logfold_col = "Log2FC_1",
  pvalue_col = "pvalue_1",
  feature_rank = "Genus",
  abundance_col = "rel_abun",
  logfold.threshold = 1.5,
  abundance.threshold = 0.001,
  label_A = "1",
  label_B = "0"
)

ggsave(
  filename = "../../../RTC/projects/lowlands_16s/results/lowlands_volcano.png",
  plot = plt,
  scaling = 1,
  width = 7,
  height = 7,
  dpi = 400
  )

data.table::fwrite(
  x = dfe$data,
  file = "../../../RTC/projects/lowlands_16s/results/lowlands_foldchange.tsv",
  sep = "\t"
)

```
## Finalize autoFlow.R
```{r}
bench::mark(
taxa <- metagenomics$new(biomData = "../../../RTC/projects/dedication/data/biom_with_taxonomy.biom",
                         metaData = "../../../RTC/projects/dedication/data/metadata.tsv",
                         treeData = "../../../RTC/projects/dedication/data/rooted_tree.newick")
)
res <- taxa$alpha_diversity(col_name = "CONTRAST_progressionT0",
                     method = "shannon",
                     paired = TRUE)

feature_contrast = c("Phylum", "Family", "Genus")
feature_filter = c("uncultured")
distance_metrics = c("unifrac")
dist_matrix = NULL
alpha_div_table = NULL
cpus = 1
feature_ranks = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

# Nested function
is_empty = function(obj) {
  if (length(obj) == 0) {
    return(NULL)
  } else {
    return(obj)
  }
}

# Plot results as list
plots <- list()

# Collect columns: CONTRAST_ and VARIABLE_
metacols <- colnames(taxa$metaData)
CONTRAST_data <- taxa$metaData[, .SD, .SDcols = grepl("CONTRAST_", metacols)]
VARIABLE_data <- taxa$metaData[, .SD, .SDcols = grepl("VARIABLE_", metacols)]


# # Ensure CONTRAST are character
# CONTRAST_colnames <- colnames(CONTRAST_data)
# taxa$metaData[, (CONTRAST_colnames) := lapply(.SD, as.character), .SDcols = CONTRAST_colnames]

#
#---------------------------------------------#
# Perform standard visualizations             #
#---------------------------------------------#
#
# CONTRAST
#
feature_nrow <- length(feature_contrast)
CONTRAST_ncol <- length(CONTRAST_data)
VARIABLE_ncol <- length(VARIABLE_data)
#
# Object manipulation
#
taxa$feature_subset(Kingdom == "Bacteria")
# Standard rank stats
plots$rankstat_plot <- taxa$rankstat(feature_ranks)

# Main loop
if (CONTRAST_ncol > 0) {

  # Load custom distance matrix if supplied
  if (!is.null(dist_matrix)) {
    dist_matrix <- read_tsv_matrix(filename = dist_matrix)
    dist_matrix <- dist_matrix[self$metaData[[self$.sample_id]], self$metaData[[self$.sample_id]]]
  }

  # Load custom rarefraction alpha diversity table if supplied
  if (!is.null(alpha_div_table)) {
    alpha_div_table <- read_rarefraction_qiime(filename = alpha_div_table)
  }

  # Initialize plot containers
  composition_plots <- matrix(list(), CONTRAST_ncol, feature_nrow)
  correlation_plots <- list()
  Log2FC_plots <- matrix(list(), CONTRAST_ncol, feature_nrow)
  alpha_div_plots <- list()
  metrics_nrow <- length(distance_metrics)
  pcoa_plots <- matrix(list(), CONTRAST_ncol, metrics_nrow)
  nmds_plots <- matrix(list(), CONTRAST_ncol, metrics_nrow)
  RDA_plots <- matrix(list(), CONTRAST_ncol, 2)

  for (i in 1:CONTRAST_ncol) {
    col_name <- colnames(CONTRAST_data)[i]
    cat(paste0("Processing ... column: ", col_name, " \n"))

    #--------------------------------------------------------------------#
    ## Alpha diversity
    #--------------------------------------------------------------------#
    if (inherits(alpha_div_table, "data.table")) {
      dt_final <- base::merge(alpha_div_table,
                              self$metaData[, .SD, .SDcols = c(self$.sample_id, col_name)],
                              by = self$.sample_id,
                              all.x = TRUE) %>%
        na.omit(cols = col_name)

       res <- diversity_plot(data = dt_final,
                             values = "alpha_div",
                             col_name = col_name,
                             palette = fetch_palette(dt_final, col_name, "Set2"),
                             method = "custom")
    } else {
      res <- tryCatch(
        {
          # Default attempt
          taxa$alpha_diversity(
            col_name = col_name,
            method = "shannon",
            paired = ifelse(!is.null(taxa$.samplepair_id), TRUE, FALSE)
          )
        },
        error = function(e) {
          cli::cli_alert_warning("alpha_diversity with paired=TRUE failed. Retrying with paired=FALSE.")
          taxa$alpha_diversity(
            col_name = col_name,
            method = "shannon",
            paired = FALSE
          )
        }
      )
    }
    
    ## Save plots and identify significant groups for composition plots & volcano plots
    alpha_div_plots[[i]] <- res$plot
    signif_pairs <- res$stats[c("group1", "group2")] #[grepl("\\ns", res$stats$p.adj.signif) ,]
    if (nrow(signif_pairs) > 0)
      conditions <- signif_pairs
      
    #--------------------------------------------------------------------#
    ## Beta diversity
    #--------------------------------------------------------------------#
    
    for (j in 1:metrics_nrow) {
      if (inherits(dist_matrix, "Matrix")) {
        res <- self$ordination(distmat = dist_matrix,
                               method = "pcoa",
                               group_by = col_name)
      } else {
        res <- taxa$ordination(metric = distance_metrics[j],
                               method = "pcoa",
                               group_by = col_name,
                               normalize = TRUE,
                               weighted = TRUE,
                               cpus = cpus)
      }
      
      ## Save plots and identify significant groups for composition plots & volcano plots
      signif_pairs <- res$anova_data[res$anova_data$p.value < 0.9, ]
      if (nrow(signif_pairs) > 0) {
        pairs_split <- strsplit(as.character(signif_pairs$pairs), " vs ")
        
        # Create group1 and group2 columns from split
        signif_pairs$group1 <- sapply(pairs_split, `[`, 1)
        signif_pairs$group2 <- sapply(pairs_split, `[`, 2)
        
        signif_pairs <- signif_pairs[c("group1", "group2")]
        
        conditions <- combine_conditions(conditions, signif_pairs)
      }
      
      pcoa_plots[[i, j]] <- patchwork::wrap_plots(res[c("scree_plot", "anova_plot", "scores_plot")],
                                                  nrow = 1) +
        plot_layout(widths = c(rep(5, 3)),
                    guides = "collect")

      # Creates temporary plot results for NMDS
      if (inherits(dist_matrix, "Matrix")) {
        res <- self$ordination(distmat = dist_matrix,
                                    method = "nmds",
                                    group_by = col_name)
      } else {
        res <- taxa$ordination(metric = distance_metrics[j],
                               method = "nmds",
                               group_by = col_name,
                               weighted = TRUE)
      }

      ## Save plots and identify significant groups for composition plots & volcano plots
      signif_pairs <- res$anova_data[res$anova_data$p.value < 0.95, ]
      if (nrow(signif_pairs) > 0) {
        pairs_split <- strsplit(as.character(signif_pairs$pairs), " vs ")
        
        # Create group1 and group2 columns from split
        signif_pairs$group1 <- sapply(pairs_split, `[`, 1)
        signif_pairs$group2 <- sapply(pairs_split, `[`, 2)
        
        signif_pairs <- signif_pairs[c("group1", "group2")]
        
        conditions <- combine_conditions(conditions, signif_pairs)
      }      

      nmds_plots[[i, j]] <- patchwork::wrap_plots(res[c("anova_plot", "scores_plot")],
                                                  nrow = 1) +
        plot_layout(widths = c(rep(5, 3)),
                    guides = "collect")
    }
    
    #--------------------------------------------------------------------#
    ## Feature composition & FOLDCHANGE
    #--------------------------------------------------------------------#
    for (j in 1:feature_nrow) {
      # Creates composition long table
      res <- self$composition(feature_rank = feature_ranks[j],
                              feature_filter = feature_filter,
                              col_name = col_name)

      # Creates composition ggplot as list
      composition_plots[[i, j]] <- composition_plot(data = res$data,
                                                    palette = res$palette,
                                                    feature_rank = feature_ranks[j],
                                                    group_by = col_name)

      # Creates correlation ggplot as list
      if (i == 1 & length(CORRELATION_data) > 0) {
        correlation_plots[[j]] <- self$correlation(feature_rank = feature_ranks[j],
                                                   feature_filter = feature_filter,
                                                   cor_columns = colnames(CORRELATION_data))
      }


      # # Creates Log2 Fold-Change (FC) ggplot as list
      # unique_groups <- unique(na.omit(RANKSTAT_data[, .SD, .SDcols = col_name]))
      # if (nrow(unique_groups) == 2) {
      #   condition_A <- unique_groups[1, ]
      #   condition_B <- unique_groups[2, ]
      #
      #   Log2FC_plots[[i, j]] <- self$differential_feature_expression(feature_rank = feature_ranks[j],
      #                                                                sample.id = "SAMPLE-ID",
      #                                                                condition.group = col_name,
      #                                                                condition_A = condition_A,
      #                                                                condition_B = condition_B,
      #                                                                feature_filter = feature_filter)[["volcano_plot"]][[1]]
      # }


    }

  }
  plots$alpha_div_plots <- is_empty(alpha_div_plots)
  plots$correlation_plots <- is_empty(correlation_plots)
  plots$composition_plots <- is_empty(composition_plots)
  plots$Log2FC_plots <- is_empty(Log2FC_plots)
  plots$pcoa_plots <- is_empty(pcoa_plots)
  plots$nmds_plots <- is_empty(nmds_plots)
  plots$rda_plots <- is_empty(RDA_plots)
}

```


## BaCo quick lookup
```{r}
taxa <- metagenomics$new(biomData = "../../../Pathology/projects/BaCo/data/WGS_baco_tissue/CAT_with_taxonomy.biom",
                          metaData = "../../../Pathology/projects/BaCo/data/metadata/metadata.tsv")

taxa$feature_glom(feature_rank = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"))

taxa <- readRDS(file = "../../../Pathology/projects/BaCo/data/WGS_baco_tissue/omicflow_aggregated.rds")

metadata <- data.table::fread("../../../Pathology/projects/BaCo/data/metadata/metadata.tsv")
taxa$metaData <- metadata
rows_to_keep <- taxa$metaData[, `SAMPLE-ID` %in% colnames(taxa$countData)]
taxa$metaData <- taxa$metaData[rows_to_keep,]

taxa$feature_subset(Kingdom == "Bacteria")
# taxa$sample_subset(group == "colitis")

plt <- taxa$ordination(metric = "bray",
                       method = "pcoa",
                       group_by = "fixative",
                       normalize = TRUE,
                       weighted = TRUE)

alpha_plt <- taxa$alpha_diversity(col_name = "side",
                                  method = "shannon",
                                  paired = FALSE)$diversity

res <- taxa$differential_feature_expression(sample.id = "SAMPLE-ID",
                                     paired.id = "PATIENT-ID",
                                     feature_rank = "Species",
                                     paired = FALSE,
                                     normalize = TRUE,
                                     condition.group = "group",
                                     condition_A = "colitis",
                                     condition_B = "control")

res$data[["p.adjust"]] <- p.adjust(res$data$pvalue, method="fdr")

volcano_plot(dt = res$data,
             X = "Log2FC_1",
             Y = "p.adjust",
             feature_rank = "Species",
             pvalue.threshold = 0.05,
             logfold.threshold = 2,
             label_A = c("colitis"),
             label_B = c("control"))

ggsave(filename = "../../../Pathology/projects/BaCo/results/biofilm_shannon.png",
       plot = alpha_plt,
       width = 10,
       height = 5,
       dpi = 400,
       scaling = 1)


ggsave(filename = "../../../Pathology/projects/BaCo/results/fixative_pcoa.png",
       plot = patchwork::wrap_plots(plt[c("anova_plot","scores_plot")], nrow = 1),
       width = 10,
       height = 5,
       dpi = 400,
       scaling = 1)

```


## Daves biom:
```{r}
taxa <- metagenomics$new(metaData = "../../../RTC/projects/MIDLOC/metadata/metadata.tsv",
                           biomData = "../../../RTC/projects/MIDLOC/preprocessed/read_annotation/metaphlan_with_taxonomy.biom")

taxa$reset()
taxa$feature_subset(Kingdom == "k__Bacteria")
taxa$normalize()

taxa$ordination(metric = "bray",
                      method = "pcoa",
                      group_by = "constipation",
                      weighted = TRUE)

ggsave(filename = "../../../RTC/projects/MIDLOC/results/pcoa_time.png",
       plot = patchwork::wrap_plots(res[c("anova_plot", "scores_plot")], nrow=1), 
       width = 8, 
       height = 5, 
       dpi = 400, 
       scaling = 1)

taxa$alpha_diversity(col_name = "Time",
                     method = "shannon")$diversity


taxa$countData

res <- taxa$differential_feature_expression(sample.id = "SAMPLE-ID",
                                            paired.id = "PATIENT-ID",
                                            feature_rank = "Genus",
                                            paired = TRUE,
                                            normalize = TRUE,
                                            condition.group = "timepoint" ,
                                            condition_A = c("prior"),
                                            condition_B = c("after"))

### Investigating relative abundance differences
significant_bacteria <- res$data[res$data[, pvalue_1 <= 0.05],][["Genus"]]

taxa$feature_glom(feature_rank = "Genus")

      final_dt <- rbind(dt[, .SD, .SDcols = !c("row_sum")],
                        dt[1:nrow(dt)][, lapply(.SD, function(x) sum(x)),
                                                                 .SDcols = !c(feature_rank, "row_sum")],
                        fill = TRUE)

res_1 <- composition_final[composition_final[, Genus %in% significant_bacteria], ]

plt_1 <- res_1 %>% 
  ggplot(mapping = aes(x = Genus,
                       y = value)) +
  geom_boxplot() +
  facet_wrap(~timepoint) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 13),
    axis.text.x = element_text(angle = 90, size = 12,
                               vjust = 0.5, hjust=1,
                               colour = "black"),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_text(size = 12, vjust=0.5),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12, colour = "black"),
    axis.text.y = element_text(colour = "black", size = 12)) +
  labs(y = "rel. abun.")


ggsave(filename = "../../../RTC/projects/MIDLOC/results/volcano_timepoint.png",
       plot = res$volcano_plot[[1]], 
       width = 8, 
       height = 5, 
       dpi = 400, 
       scaling = 1)


### Two-way ANOVE with repeated measures
counts <- data.table::data.table(as.matrix(taxa$countData))
long_counts <- data.table::melt(
  counts,
  measure.vars = data.table::patterns("^MIDLOC"),
  variable.name = "sample",
  value.name = "counts"
)

long_counts[, c("patient", "timepoint") := data.table::tstrsplit(sample, "T", perl=TRUE)]
final_dt <- merge()

## Testing for normality
final_dt$patient <- as.factor(final_dt$patient)
final_dt$timepoint <- as.factor(final_dt$timepoint)

library(stats)

anova_res <- aov(div ~ timepoint + Error(patient/timepoint), data = div)

summary(anova_res)


# Instead of looking at relative abundance, what about alpha diversity ?
res <- taxa$alpha_diversity(col_name = "Time",
                            method = "shannon")

# Fit generalized linear mixed model
M1 <- glmmTMB::glmmTMB(
  counts ~ timepoint + (1|patient),
  ziformula = ~ timepoint,
  family = glmmTMB::beta_family(),
  data = final_dt
)

# Fit generalized linear mixed model using Bayesian
fit <- brms::brm(
  brms::bf(counts ~ timepoint + (1 | patient)), 
  family = brms::zero_inflated_beta(), 
  data = final_dt,
  chains = 4, iter = 5000, warmup = 1000, seed = 4445
)


confint(M1, parm = "beta_")

## QQ plot for mixed models
# Generate simulated residuals
sim_residuals <- DHARMa::simulateResiduals(M1)
plot(sim_residuals)

# Fit linear mixed-effects model
M2 <- lme4::lmer(div ~ timepoint + (1|patient),
                 data = div,
                 REML = FALSE)

# test for singularity:
lme4::isSingular(M2)

# Diagnostic plot
## residuals
plot(M2)

## summary 
summary(M2)

## QQ plot base R
qqnorm(resid(M2))
qqline(resid(M2))

## QQ plot for mixed models
# Generate simulated residuals
sim_residuals <- DHARMa::simulateResiduals(M2)
plot(sim_residuals)

anova(M2)


```

## Test case
```{r}
taxa <- metagenomics$new(metaData = "../../../Pathology/projects/PRELEDE/data/metagenomics/metadata.tsv",
                           biomData = "../../../Pathology/projects/PRELEDE/data/metagenomics/CAT_with_taxonomy.biom")

colnames(taxa$featureData) <- c("ID", "feature_id", "probability", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

taxa$rankstat()

taxa$feature_subset(Kingdom == "Bacteria")

res <- taxa$alpha_diversity(col_name = "tissue_type",
                     method = "shannon")

res <- taxa$ordination(metric = "bray",
                method = "pcoa",
                group_by = "tissue_type")


```

# Benchmark of sub-sampling
```{r}
taxa <- metataxonomics$new(metaData = "../../../RTC/projects/NGID_16S/data/metadata.tsv",
                           biomData = "../../../RTC/projects/NGID_16S/data/MERGE-CP1393-CP1394_curated_table.biom")

taxa$feature_subset(Kingdom == "Bacteria")

contrast <- colnames(taxa$metaData)[5]

res <- taxa$triplot(feature_rank = "Genus",
                    feature_filter = NA,
                    metadata.col = contrast,
                    pairwise = FALSE,
                    choice_dim = c("RDA1", "PC1"))

vegan::permutest(res$model, permutations = 999, pairwise = FALSE)


distmat <- data.table::fread("../../../Pathology/projects/HEREDITARY/novaseq_benchmark/subsamples/weighted-unifrac-100k-base.tsv",
                             header = TRUE)
mat <- as.matrix(distmat[,.SD, .SDcols = !c("V1")])

alpha_div <- taxa$alpha_diversity(col_name = "type", method = "shannon")
beta_div <- taxa$ordination(metric = "unifrac", 
                            method = "pcoa",
                            group_by = "type",
                            distmat = as.dist(mat))

main_panel <- patchwork::wrap_plots(alpha_div, nrow=1) / (patchwork::wrap_plots(beta_div, nrow=1))

ggsave(filename = "../../../Pathology/projects/HEREDITARY/novaseq_benchmark/subsamples/results/100k_base.png",
       plot = main_panel,
       dpi = 400,
       scale = 0.9,
       width = 30,
       height = 20,
       units = "cm")


# Splitting biom by sample in tsv files
taxa$feature_subset(Kingdom == "Bacteria")
taxa$sample_subset(study_group == "HBS")

taxa$feature_glom(feature_rank = "Genus")
taxa$normalize()

dimnames(taxa$countData)[[1]] <- taxa$featureData$Genus
col_labels <- colnames(taxa$countData)

for (i in 1:ncol(taxa$countData)) {
  column_data <- as.data.frame(taxa$countData[, i], row.names = rownames(taxa$countData))
  colnames(column_data) <- col_labels[i]
  write.csv(column_data, 
            file = paste0("../../../Pathology/projects/HEREDITARY/data/HBS/samples_genus/", col_labels[i], ".csv"), 
            row.names = TRUE)
}

```

# HEREDITARY Exploration Data Analysis
```{r}
taxa <- metataxonomics$new(metaData = "../../../Pathology/projects/HEREDITARY/data/metadata.tsv",
                           biomData = "../../../Pathology/projects/HEREDITARY/data/biom_with_taxonomy_mod4.biom",
                           treeData = "../../../Pathology/projects/HEREDITARY/data/rooted_tree_mod4.newick")
taxa$reset()

taxa$feature_subset(Kingdom == "Bacteria")

read_stats <- data.table::fread("../../../Pathology/projects/HEREDITARY/data/read_stats.tsv")

read_stats_perc <- read_stats[, .SD, .SDcols = c("input", "trimmed%", "assembled%", "denoised%", "filtered%", "nonchim%")]
read_stats_perc$input <- 100

read_stats_long <- data.table::melt(data = read_stats_perc,
                                   measure.vars = colnames(read_stats_perc),
                                   variable.name = "conditions",
                                   variable.factor = FALSE,
                                   value.name = "values")

read_stats_long[, conditions := factor(conditions, levels = c("input", "trimmed%", "assembled%", "filtered%", "denoised%", "nonchim%"))]



read_stats_long %>% 
  ggplot(mapping = aes(x = conditions,
                       y = values)) +
  geom_boxplot() +
  


```

# PCoA
```{r}
distmat <- data.table::fread("../../../RTC/projects/dedication/data/distance_matrices/weighted_unifrac_distance_matrix.qza.txt",
                             header = TRUE)

rownames <- distmat[[1]]
colnames <- names(distmat)[-1]
values <- as.matrix(distmat[, -1, with = FALSE])

sparse_matrix <- sparseMatrix(
  i = rep(1:nrow(values), ncol(values)),
  j = rep(1:ncol(values), each = nrow(values)),
  x = as.vector(values)
)
rownames(sparse_matrix) <- rownames
colnames(sparse_matrix) <- colnames

# Subset by study_group
taxa$reset()
taxa$feature_subset(Kingdom == "Bacteria")

# Subset distance matrix
subset_mat <- sparse_matrix[taxa$metaData[["SAMPLE-ID"]], taxa$metaData[["SAMPLE-ID"]]]

pcoa <- taxa$ordination(distmat = subset_mat, 
                        method = "pcoa",
                        group_by = "RANKSTAT_inclusionsite")

ggsave(filename = "../../../Pathology/projects/HEREDITARY/results/BaCo_pcoa_mod4.png",
       plot = patchwork::wrap_plots(pcoa, nrow=1),
       units = "cm",
       width = 20,
       height = 7,
       limitsize = FALSE,
       dpi = 400,
       scaling = 0.5)

pcoa

```

# Shannon
```{r}
df_shannon <- data.table::fread("../../../Pathology/projects/HEREDITARY/data/shannon_mod4.csv")

# Pivot into long table
shannon_long <- data.table::melt(data = df_shannon,
                                 measure.vars = colnames(df_shannon)[grepl("depth-", colnames(df_shannon))], 
                                 variable.name = "iters",
                                 variable.factor = FALSE,
                                 value.name = "alpha_div")
# Corrects colnames
colnames(shannon_long) <- c("SAMPLE-ID", "iters", "alpha_div")
# Adds new column
shannon_final <- base::merge(shannon_long, 
                             taxa$metaData[, .SD, .SDcols = c("SAMPLE-ID", "type", "study_group")], 
                             by = "SAMPLE-ID", 
                             all.x = TRUE)

HBS_rows <- shannon_final[, study_group == "HBS"]
HBS <- shannon_final[HBS_rows, ] 

BaCo_rows <- shannon_final[, study_group == "BaCo"]
BaCo <- shannon_final[BaCo_rows, ]

plt2 <- diversity_plot(dt = HBS,
                      values = "alpha_div",
                      col_name = "type",
                      palette = fetch_palette(HBS, "type", "Set2"),
                      method = "shannon")

ggsave(filename = "../../../Pathology/projects/HEREDITARY/results/HBS_shannon_mod4.png",
       plot = plt2,
       units = "cm",
       width = 10,
       height = 7,
       limitsize = FALSE,
       dpi = 400,
       scaling = 0.5)

plt2
plt1

```

## Volcano plots
```{r}
taxa$reset()
taxa$feature_subset(Kingdom == "Bacteria")

plt1 <- taxa$differential_feature_expression(feature_rank = "Genus",
                                             sample.id = "SAMPLE-ID",
                                             condition.group = "RANKSTAT_smoking",
                                             condition_A ="quit",
                                             condition_B = "active",
                                             feature_filter = c("uncultured", "unidentified", "metagenome"),
                                             pvalue.threshold = 0.05)

ggsave(filename = "../../../RTC/projects/dedication/result/volcano_smoking.png",
       plot = plt1$volcano_plot[[1]],
       width = 7,
       height = 5,
       limitsize = FALSE,
       dpi = 400,
       scaling = 1)
plt2
```

## PANEL
```{r}
main_panel_1 <- (wrap_elements(plt_1) + 
                   wrap_elements(plt_2) +
                   plot_layout(widths = c(5, 6))) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = 'bold', size = 15))

ggsave(filename = "../../../Pathology/projects/HEREDITARY/results/studygroups_composition.png",
       plot = main_panel_1,
       units = "cm",
       width = 20,
       height = 10,
       limitsize = FALSE,
       dpi = 400,
       scaling = 0.5)

```

## Assesing confounders in (longitudional) sample data
```{r}
# library("tmle") # for non longitudional data
library("ltmle")

data(sampleDataForLtmleMSM)
Anodes <- grep("^A", names(sampleDataForLtmleMSM$data))
Lnodes <- c("CD4_1", "CD4_2")
Ynodes <- grep("^Y", names(sampleDataForLtmleMSM$data))
msm.weights <- matrix(1:12, nrow=4, ncol=3) #just an example (can also use a 200x3x4 array),
#or NULL (for no weights), or "empirical" (the default)
result2 <- ltmleMSM(
  sampleDataForLtmleMSM$data, 
  Anodes=Anodes,
  Lnodes=Lnodes, 
  Ynodes=Ynodes,
  survivalOutcome=TRUE,
  regimes=sampleDataForLtmleMSM$regimes,
  summary.measures=sampleDataForLtmleMSM$summary.measures, 
  final.Ynodes=Ynodes,
  working.msm="Y ~ male + time + I(pmax(time - switch.time, 0))",
  msm.weights="empirical", 
  estimate.time=FALSE)


summary(result2, )


library("tmle")
n <- 250
W <- matrix(rnorm(n*3), ncol=3)
colnames(W) <- paste("W",1:3, sep="")
A <- rbinom(n,1, plogis(0.6*W[,1] +0.4*W[,2] + 0.5*W[,3]))
Y <- rbinom(n,1, plogis(A + 0.2*W[,1] + 0.1*W[,2] + 0.2*W[,3]^2 ))

result2 <- tmle(Y,A,W, family="binomial", Qform="Y~A+W1+W2+W3", gform="A~W1+W2+W3")

summary(result2)


```

## Causal inference: Generalized propensity scores (GPS)
```{r}
model <- lm(T ~ X1 + X2 + X3, data = yourdata)

# Predicted mean for each subject
mu_hat <- predict(model)
# Residual standard deviation
sigma_hat <- summary(model)$sigma
# For each subject: GPS = density at their observed treatment value
library(stats)
yourdata$gps <- dnorm(yourdata$T, mean = mu_hat, sd = sigma_hat)

# Video tutorial: https://www.practicalpropensityscore.com/continuous.html

```


## Causal inference, discovery of confounding effects
```{r}
## Binary treatment T -> Y with confounder covariate matrix W -> T and W -> Y
library(DoubleML)

# Prepare data: X (confounders), y (outcome), d (treatment)
dml_data <- DoubleMLData$new(data, y_col = "Y", d_cols = "T", x_cols = confounders)

# Specify learner models for nuisance functions
learner <- mlr3::lrn("regr.ranger")  # random forest regression
learner_binary <- mlr3::lrn("classif.ranger", predict_type = "prob")  # for PS if binary

# Initialize DoubleML PLR object (Partial Linear Regression - binary treatment)
dml_plr <- DoubleMLPLR$new(dml_data, ml_g = learner, ml_m = learner_binary)

# Fit model and get treatment effect
dml_plr$fit()
dml_plr$coef


## Longitudional data, M1 -> M2 -> M3, where (candidate) confounder covariate matrix W -> M1, W -> M2, W -> M3.
library(DoubleML)

# Example at time t=1
dml_data_t1 <- DoubleMLData$new(data, y_col = "M2", d_cols = "M1", x_cols = c("Age", "Sex", W0_covariates))

# Fit DML model for effect of M1 on M2
ml_l <- mlr3::lrn("regr.ranger")
ml_m <- mlr3::lrn("regr.ranger")
dml_plr_t1 <- DoubleMLPLR$new(dml_data_t1, ml_l = ml_l, ml_m = ml_m)
dml_plr_t1$fit()
print(dml_plr_t1)

# Repeat for (M2 → M3), controlling for prior treatments and covariates

```

# (stepwise) Regression models
```{r}
both_model <- lm(mpg ~ ., data=mtcars)

final_model <- step(both_model, 
                    direction="both", 
                    trace = 0)

# computing AIC and BIC for a model using:
stats::AIC()
stats::BIC()

# models
lm # linear regression
glm # logistic regression

# Detect multicollinearity factorrs with:
regclass:VIF()

null_model = lm(Fuel ~ 1, data=fuel2001_4)
full_model = lm(Fuel ~ Tax + Dlic + Income_2 + log_Miles + Tax:Dlic + Tax:Income_2 + Tax:log_Miles + Dlic:Income_2 + Dlic:log_Miles + Income_2:log_Miles + I(Tax^2) + I(Tax^3) + I(Tax^4) + I(Dlic^2) + I(Dlic^3) + I(Dlic^4) + I(Income_2^2) + I(Income_2^3) + I(Income_2^4) + I(log_Miles^2) + I(log_Miles^3) + I(log_Miles^4), data=fuel2001_4)

# step function
model_step = step(model, 
                  scope = list(lower=null_model, upper=full_model),
                  direction = "both", k= 2) 
                  # direction = c("both", "backward", "forward")
                  # k=2 for AIC; k=log(nrow(train_data)) for BIC


```
