---
title: "Metadata File Specification"
description: "Specification for OmicFlow sample metadata files."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metadata File Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

Every class in OmicFlow starts with metadata validation, which checks that sample IDs match those in the abundance tables.  
The validation is defined via a JSON format in the abstract class `omics`, it takes as input a **CSV/TSV** file or a `data.table`.  
In both cases, the sample metadata must contain a header (first line if you supply a file) where **each row = one sample**.  
Additional column names not mentioned here are allowed and will be ignored during metadata validation.

Below are the exact specifications required for the metadata structure.

## **Minimum requirement**
- **`SAMPLE_ID`** ➡ every row **must** have a unique, non‑empty sample identifier.
- No spaces are allowed in IDs — use underscores `_` or dashes `-` instead.

Example:

| SAMPLE_ID | SAMPLEPAIR_ID | CONTRAST_Treatment | VARIABLE_Age |
|-----------|---------------|--------------------|--------------|
| S1        | P1            | Drug               | 42           |
| S2        | P1            | Placebo            | 36           |
| S3        | P2            | Drug               | 51           |

---

## **Column types and naming rules**

### 🔹 Required column
| Column       | Type    | Rules                                                |
|--------------|---------|------------------------------------------------------|
| `SAMPLE_ID`  | string  | Unique, no spaces, one per sample row                |

### 🔹 Optional standard columns
| Column         | Type    | Rules                                                               |
|----------------|---------|---------------------------------------------------------------------|
| `SAMPLEPAIR_ID`| string  | Optional — no spaces. Use when samples are paired and belong to an individual source/subject |

### 🔹 Pattern‑based columns
You can define extra variables using special prefixes:
- **`CONTRAST_...`** → grouping/category labels used in differential comparisons  
  Example: `CONTRAST_Treatment` with values `Drug` / `Placebo`
- **`VARIABLE_...`** → numeric or string variables for statistical analysis  
  Example: `VARIABLE_Age` with values `42`, `51`, etc.

The pattern-based columns are only used during the `autoFlow` function. At the moment only columns with prefix `CONTRAST_` are supported.

## Examples

You can always check metadata up‑front before loading any other data files by creating an `omics` object with only the `metaData` argument:

```{r setup}
library(OmicFlow)

# Check from filepath
metadata_file <- system.file("extdata", "metadata.tsv", package = "OmicFlow")
test <- omics$new(metaData = metadata_file)

# Check from `data.table` object
metadata <- data.table::data.table("SAMPLE_ID" = paste0("Sample_", 1:5))
test <- omics$new(metaData = metadata)
```
