---
title: "Visualisations & Statistics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{visualisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

Here we give some examples in how to use the `metagenomics` class from loading your data to visualization and some statistical testing. In the study of microbial communities we often look at the alpha- (i.e. `alpha_diversity` function), beta- (i.e. `ordination` function) diversities, and compositions and or fold-change differences.

The following functions do not change the object itself and always output the results.

- `rankstat`
- `alpha_diversity`
- `ordination`
- `composition`
- `foldchange`

Other similar functions that are not discussed:

- `distance` - [docs](https://agusinac.github.io/OmicFlow/reference/omics.html#omics-distance-)
- `autoFlow` - [docs](https://agusinac.github.io/OmicFlow/reference/omics.html#omics-autoflow-)

## Loading and data-wrangling

Below an example is shown for creating a `metagenomics` class from text files. The same procedure is also possible via `biomData` see the [metagenomics$new](https://agusinac.github.io/OmicFlow/reference/metagenomics.html#metagenomics-new-), and the loaded files from text or BIOM can also be written out as a `BIOM` formatted hdf5 file via [write_biom](https://agusinac.github.io/OmicFlow/reference/metagenomics.html#method-metagenomics-write_biom).

```{r}
library("OmicFlow")
library("ggplot2")

## Loading example set
metadata_file <- system.file("extdata", "metadata.tsv", package = "OmicFlow")
counts_file <- system.file("extdata", "counts.tsv", package = "OmicFlow")
features_file <- system.file("extdata", "features.tsv", package = "OmicFlow")
tree_file <- system.file("extdata", "tree.newick", package = "OmicFlow")

## Initializing taxa object
taxa <- metagenomics$new(
  metaData = metadata_file,
  countData = counts_file,
  featureData = features_file,
  treeData = tree_file
)

## Subseting to `Bacteria`
taxa$feature_subset(Kingdom == "Bacteria")

## Classical total-sum scaling by sample-level
taxa$scale(method = "tss")
```

## Number of classified ASVs
```{r}
plt <- taxa$rankstat(feature_ranks = c("Kingdom", "Phylum", "Family", "Genus", "Species"))
plt
```

Please visit [rankstat](https://agusinac.github.io/OmicFlow/reference/omics.html#omics-rankstat-) for more information.

## Alpha- and Beta-diversity

### Alpha diversity

```{r}
alpha_res <- taxa$alpha_diversity(
  col_name = "treatment",
  metric = "shannon",
  paired = FALSE # If TRUE it performs wilcox signed rank test
)
alpha_res$plot
```

Please visit [alpha-diversity docs](https://agusinac.github.io/OmicFlow/reference/omics.html#omics-alpha-diversity-) for more information.

### Beta diversity

By default PERMANOVA is applied pairwise against each group within the specified contrast, via `group_by` that is used in `pairwise_adonis`. The permutation design in `vegan::adonis2` is by default set to `free`. But this may not always be the right test when you have paired samples and you also want to restrict permutations between correlated values. Therefore, `pairwise_adonis` supports a custom permutation design, which can be constructed via [permute](https://cran.r-project.org/web/packages/permute/vignettes/permutations.html) and fed into `vegan::adonis2` as a function via `pairwise_adonis` with the flag `perm_design`.

Performing ordination with tests without a permutation design
```{r}
set.seed(1970)

# Perform ordinations with in-built distance matrix computation
#--------------------------------------------------------------------------------
beta_div <- taxa$ordination(
    metric = "unifrac",
    method = "pcoa",
    group_by = "treatment",
    perm = 999
)

patchwork::wrap_plots(
    beta_div[c("scree_plot", "anova_plot", "scores_plot")],
    nrow = 1)
```

Beta-diversity with a custom dissimilarity matrix
```{r, eval = FALSE}
# Add a custom pre-computed distance matrix (e.g. from QIIME2)
#--------------------------------------------------------------------------------
qiime_unifrac <- data.table::fread("weighted-unifrac-matrix.tsv", header=TRUE)
distmat <- Matrix::Matrix(as.matrix(qiime_unifrac[, .SD, .SDcols = !c("V1")]))
rownames(distmat) <- colnames(distmat)
distmat <- distmat[taxa$metaData[["SAMPLE_ID"]], taxa$metaData[["SAMPLE_ID"]]]
distmat <- as.dist(distmat) 

beta_div <- taxa$ordination(
    distmat = distmat,
    method = "pcoa",
    group_by = "treatment",
    perm = 999
)

```

Beta-diversity with a permutation design, you can also run `pairwise_adonis` outside of the classes, check the [docs](https://agusinac.github.io/OmicFlow/reference/pairwise_adonis.html).
```{r, eval = FALSE}
# Add a custom permutation design via `perm_design`
#--------------------------------------------------------------------------------
## taxa$ordination() automatically will input taxa$metaData inside the supplied function.
perm_design_func <- function(meta) {
  base::with(
    data = meta,
    expr = permute::how(
      nperm = 999,
      plots = permute::Plots(meta$SAMPLEPAIR_ID, type = "none"), # In case samplepair ids is supplied
      within = permute::Within(type = "free")
    )
  )
}

beta_div <- taxa$ordination(
    metric = "unifrac",
    method = "pcoa",
    group_by = "treatment",
    perm_design = perm_design_func
)
```

Please visit [beta-diversity docs](https://agusinac.github.io/OmicFlow/reference/omics.html#omics-ordination-) for more information

## Compositions of bacteria
Compositional plots can be made from stacks or over all samples, the `feature_rank` uses [feature_merge](https://agusinac.github.io/OmicFlow/reference/omics.html#omics-feature-merge-) in the background but it doesn't modify the object.
```{r}
## Plotting only Phylum rank
phylum <- taxa$composition(
  feature_rank = "Phylum",
  
  # This can be a vector of multiple keywords or `NULL`
  feature_filter = c("uncultured"),
  
  # The maximum is 15 for visualisation purposes.
  feature_top = 15,
  
  # This can be a vector of multiple keywords
  col_name = "CONTRAST_sex"
)

composition_plot(
  data = phylum$data,
  palette = phylum$palette,
  feature_rank = "Phylum",
  group_by = "CONTRAST_sex"
)
taxa

## Plotting only Genus rank
genera <- taxa$composition(
  feature_rank = "Genus",
  feature_filter = c("uncultured"),
  feature_top = 15,
  col_name = "CONTRAST_sex"
)

## Plotting genera grouped by samples
composition_plot(
  data = genera$data,
  palette = genera$palette,
  feature_rank = "Genus",
  group_by = "CONTRAST_sex"
)

## Plotting only genera over all samples
composition_plot(
  data = genera$data,
  palette = genera$palette,
  feature_rank = "Genus"
)

```

## Fold-change
The `foldchange` of the `metagenomics` class differs from those in `omics` and `proteomics`. Namely, it expects the data to be proportional, and it cannot handle log-transformed data. **This will soon change!**
```{r, eval = FALSE}
taxa$foldchange(
  feature_rank = "Genus",
  paired = FALSE,
  condition.group = "treatment",
  
  # Able to split your data into male/female
  group_by = "CONTRAST_sex",
  condition_A = c("healthy"),
  condition_B = c("tumor")
)
```

Please visit [foldchange](https://agusinac.github.io/OmicFlow/reference/metagenomics.html#metagenomics-foldchange-) for more information.

# Proteomics

In the case of `proteomics` or other `omics` classes, not all functions discussed here can be applied. For example in the context of proteomics, data is often log-transformed and methods such as unifrac or alpha-diversity metrics cannot be applied on data with negative values. On top of that, visualising compositions or `rankstat` isn't always informative. Therefore, other omics data often follow the log-transformed route, where `ordination` with the `euclidean` distance can be applied to investigate differences in communities, followed by fold-change differences.